A Simple Minibatch Acceptance Test for MCMC

### 2016-05-16 nips_2016.tex
updates to abstract

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^np(x_i \mid \theta)$ of
parameter $\theta$ based on conditionally independent data
$\{x_i\}_{i=1}^n$ is the target distribution which one aims to
approximate by sampling $\theta$ values. When $n$ is large (e.g., in
the billions), it is computationally difficult to compute the exact
posterior because the likelihood must be evaluated on each $x_i$.

Recent success in machine learning optimization in the presence of large datasets have been in part
due to the ubiquitous use of stochastic gradient methods, where a subset of the data is used each
iteration to compute a noisy estimate of the gradient. These Robbins-Munro
algorithms~\cite{RobbinsMonro1951} have historically not been used with MCMC methods. A recent
result from~\cite{cutting_mh_2014} attempted to bridge the gap between minibatch optimization and
Bayesian inference by proposing an \emph{adaptive minibatch} method. Their method uses a subset of
the $N$ data points for the MH test each iteration, but requires incrementally increasing the
minibatch size so that it is more confident in its prediction on accepting or rejecting.

We argue for an alternative approach to using MCMC methods in the context of big data problems. In
general, model optimization involves finding one point estimate of optimal parameters based on a
user's goals, which typically refer to MAP or ML estimates. In addition, the more data one has, the
sharper the posterior distribution's modes. In the extreme case, the posterior would collapse to the
points corresponding to the modes, and it would be difficult for MCMC methods to escape.

Therefore, we are not interested in exact inference of the posterior model parameters for a very
large dataset, but rather, we want to perform either MAP or ML estimation. In both cases the goal is
to find a good posterior mode, but the exact form of the distribution is arbitrary. In particular,
non-unit powers of the distribution are fine. That corresponds to changing the temperature of the
distribution to make it ``flatter''.  Rather than fixing the posterior temperature and trying to use
adjustable minibatch size to achieve it (which ends up reducing to nearly-full-batch updates) we
propose to run at the natural temperature induced by a given minibatch size. We can always increase
the minibatch size to reduce the temperature, but there is no need to do this beyond the point where
the posterior mode is sharp and well-defined. That will occur at some fixed number of points and far
less than the full dataset.

To this end, we present a new Metropolis-Hastings test that can fit in this framework of minibatch
MCMC methods. We present theoretical and experimental results on our method and show that it can be
the method of choice for minibatch MCMC. The contributions of this paper are as follows:

\begin{enumerate}[noitemsep]
    \item We develop a new minibatch MH test which satisfies detailed balance.
    \item Our MH test is more efficient than the adaptive sampling method.
    \item The test can incorporate temperature considerations to control mixing rates.
\end{enumerate}

{\color{blue}
Daniel: This section needs work in describing our argument carefully. I will decrease the focus on
temperature and increase focus on the new MH test once the results are clearer.
}

>>>>

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^np(x_i \mid \theta)$ of
parameter $\theta$ based on conditionally independent data
$\{x_i\}_{i=1}^n$ is the target distribution which one aims to
approximate by sampling $\theta$ values. When $n$ is large, it is
very expensive to compute the exact posterior because the
likelihood must be evaluated on every $x_i$.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.may
Minibatch updates produce many improvements in the model for each pass
over the dataset, and have high sample efficiency. They also map very
well onto hardware such as GPUs. MCMC methods in contrast, typically
require calculations over the full dataset to produce a new sample. A
recent result from~\cite{cutting_mh_2014} attempted to bridge the gap
between minibatch optimization and MCMC by proposing an \emph{adaptive
  minibatch} method. Their method uses a subset of the $N$ data points
for the MH test each iteration. However the number of points is not
known a priori, and is adjusted dynamically during every test. This
undermines many of the practical advantages of minibatch calculations
- GPU kernels work best for certain minibatch sizes, memory blocks
can be reused when minibatch size is fixed etc.

Here we develop an alternative minibatch acceptance test for MCMC
methods which uses only one fixed-size minibatch per test. Instead of
dynamically adjusting the batch size, we add a prequisite for the test
based on the variance of the log acceptance probability. If this
variance is small enough, the test requires only a single step. This
variance changes slowly on typical problems, and we can estimate it
before performaning the accpeptance test. The minibatch variance
descreases with the minibatch size, so the variance measurement can be
used to set the minimum minibatch size.

There is an additional advantage when using the test for Bayesian
posterior inference. As dataset size grows, the posterior distribution
sharpens and will often concentrate to a few sharp peaks. Samples from
this distribution will come from those peaks, and will easily become
stuck in one of them (see section 5.1).  The samples are not exploring
the posterior parameter space, and are unlikely to concentrate at a
strong local optimum. They therefore fail to accurately represent a
sample of the posterior because of inadequate search (inadequate
mixing). We argue that it is much more natural to sample the
distribution at a different (higher) temperature determined by the
minibatch size rather than the dataset size. This temperature is
determined in a natural way by the variance prerequisite for our
test. Starting at a high temperature (small minibatch size) which
mixes easily, one can reduce the temperature (increasing minibatch
size) until one reaches the desired target distribution temperature.
Although this approach also involves changing minibath size, the
changes increase minibatch size monotonically, and only a few changes
are performed in the course of a sampling session.
Annealing can also be used for MAP and ML estimation because changing temperature preserves
posterior modes.

The contributions of this paper are as follows:

\begin{enumerate}[noitemsep]
    \item We develop a new minibatch acceptance test which satisfies detailed balance.
    \item We compare performance of our new test and the adaptive sampling method.
    \item Experiments using the approach for posterior estimation.
\end{enumerate}

%{\color{blue}
%Daniel: This section needs work in describing our argument carefully. I will decrease the focus on
%temperature and increase focus on the new MH test once the results are clearer.
%}
